Parsing Postfix log files

### 2008-10-30 doc/logparser-acronyms.tex
Fixed underfull hbox warnings. Dealt with more corrections; nearly finished (if I ignore the 103 XXX notes).

\newacronym{SQL}{Structured Query Language}{description={
    SQL is the standard language used for database querying, modification
    and maintenance.  Some information about SQL, including its history,
    can be found at~\cite{Wikipedia-sql}; a good introduction can be found
    at~\cite{sql-for-web-nerds}, others are~\cite{w3schools-sql-tutorial,
    sqlcourse.com}.
}}

@@

\newacronym{HTML}{Hypertext Markup Language}{description={
    HTML is the markup language designed for writing web
    pages~\cite{Wikipedia-html,w3schools-html}.
}}

>>>>

\newacronym{SQL}{Structured Query Language}{description={
    SQL is the standard language used for database querying, modification
    and maintenance.  Some information about SQL, including its history,
    can be found at \url{http://en.wikipedia.org/wiki/SQL}; a good
    introduction can be found at \url{http://philip.greenspun.com/sql/}.
}}

@@

### 2008-10-30 doc/logparser-background.tex
Fixed underfull hbox warnings. Dealt with more corrections; nearly finished (if I ignore the 103 XXX notes).

            \item New columns can be added to the tables used by the
                program, with sufficient DEFAULT clauses or a clever
                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
                for explanations of these terms,
                e.g.~\cite{sql-for-web-nerds}}

@@

    \item \SQL{} is reasonably standard and many people will already be
        familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn (a good
        introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
        others are~\cite{w3schools-sql-tutorial,sqlcourse.com}).  Although
        every vendor implements a different dialect of \SQL{}, the basics
        are the same everywhere (analogous to the overall similarities and
        minor differences between Irish English, British English, American
        English, and Australian English).  Depending on the database in use
        there may be tools available that reduce or remove the requirement
        to know \SQL{}; several are available for SQLite
        (\url{http://www.sqlite.org/}), the default database used by the
        parser implementation.

@@

    \caption{Example attributes sent to policy servers}\label{Example
    attributes sent to policy servers}

    XXX HOW CAN I CENTRE THE TABLE\@?
    \begin{tabular}[]{ll}

>>>>

            \item New columns can be added to the tables used by the
                program, with sufficient DEFAULT clauses or a clever
                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
                for explanations of these terms, e.g.\
                \url{http://philip.greenspun.com/sql/}.}

@@

    \item \SQL{} is reasonably standard and many people will already be
        familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn (a good
        introduction to \SQL{} can be found at
        \url{http://philip.greenspun.com/sql/},
        others are \url{http://www.w3schools.com/sql/default.asp},
        \url{http://sqlcourse.com/}).  Although
        every vendor implements a different dialect of \SQL{}, the basics
        are the same everywhere (analogous to the overall similarities and
        minor differences between Irish English, British English, American
        English, and Australian English).  Depending on the database in use
        there may be tools available that reduce or remove the requirement
        to know \SQL{}; several are available for SQLite
        (\url{http://www.sqlite.org/}), the default database used by the
        parser implementation.

@@

    \caption{Example attributes sent to policy servers}
    \empty{}\label{Example attributes sent to policy servers}

    XXX HOW CAN I CENTRE THE TABLE\@?\hfill{}

    \begin{tabular}[]{ll}

### 2008-10-30 doc/logparser-conclusion.tex
Fixed underfull hbox warnings. Dealt with more corrections; nearly finished (if I ignore the 103 XXX notes).

@@

Given the diverse nature of the log lines and the ease with which
administrators can cause new lines to be logged (\sectionref{postfix
background}), enabling users to easily extend the parser to deal with new
log lines is a design imperative (\sectionref{parser design}).  Despite the
resulting initial increase in complexity the task is quite tractable,
though it does raise efficiency and optimisation questions (answered in
\sectionref{parser efficiency}).  Providing a tool (\sectionref{creating
new rules}) to ease the generation of \regexes{} from unparsed log lines
should greatly help users add rules to parse formerly unparsed log lines.

The implementation not only substantially eases the parsing of new log
lines, it makes adding new actions (\sectionref{adding new actions}) a
relatively simple task.  The simplicity of adding a new action frees the
implementor from worrying about how their action might disrupt parsing of
other log lines or the behaviour of other actions, allowing their
concentration to focus on correctly implementing their new action.

The division of the parser into rules, actions and framework is unusual
because rules are separated so completely from the actions and framework;
although parsers are often divided (parsers based on the combination of
\texttt{lex} and \texttt{yacc}~\cite{lex-and-yacc} 
being an obvious example), the parts are usually quite internally
interdependent, and will be combined into a complete parser by the
compilation process; in contrast \parsername{} keeps the rules and actions
separate until the parser runs.  The actions and framework are not as
completely separated, as the actions depend on services provided by the
framework; however the actions and framework are not tightly integrated: it
would be possible, with some work, to completely separate the two.

The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
behaviour}) exhibited by the rules and actions is also interesting, and is
discussed after the flow chart (\sectionref{flow-chart}) and explanation of
the paths through the parser.  This emergent behaviour greatly eases the
process of adding new actions (\sectionref{adding new actions}), as the
actions do not have to be inserted into an explicit flow of control; new
actions will naturally find their place in the paths through the parser.

The flow of control in this parser is quite different from that of most
other parsers.  In most traditional parsers the parser has a current state,
and each state has a fixed set of acceptable next states, with unacceptable
states causing parsing to fail --- e.g.\ when a \textbf{C} parser sees the
keyword \textbf{for} it expects to immediately see a left parenthesis, with
any other input causing parsing to fail.\footnote{Comments will already
have been removed by the preprocessor.}  This parser is different: the
rule which matches the next log line for a mail dictates the action that
will be invoked, which is equivalent to the next state in other parsers;
thus the next log line for a mail dictates the next state for that mail.

The real difficulties arise once the parser is successfully dealing with
90\% of the log lines, as the irregularities and complications previously
explained only begin to become apparent once the vast majority of log lines
have been parsed successfully.  Adding new rules to deal with numerous,
infrequently occurring log lines is a simple task, albeit tiresome (though
alleviated by the tool described in \sectionref{adding new actions}),
whereas dealing with mails where information appears to be missing is much
more grueling.  Trawling through the log files, looking for something out
of the ordinary, possibly hundreds or even thousands of lines away from the
last mention of the queueid in question, is extremely time consuming and
error prone.  Sometimes the task is not to spot the line that is unusual,
but to spot that a line normally present is missing, i.e.\ to realise that
one line amongst hundreds is absent.  In all cases the evidence must be
used to construct a hypothesis to explain the irregularities, and that
hypothesis must then be tested in the parser; if successful, the parser
must be modified to deal with the irregularities, without adversely
affecting the existing functionality.  The complications described in
\sectionref{complications} were solved in the order they are
described in, and that order closely resembles the frequency with which
they occur; the most frequently occurring complications dominate the
warning messages produced, and so naturally they are the first
complications to be dealt with.

>>>>

XXX THIS NEEDS TO BE EXTENDED\@.

XXX THERE IS NO MENTION OF RESULTS/EFFICIENCY\@.

@@

Given the diverse nature of the log lines and the ease with which
administrators can cause new lines to be logged (\sectionref{postfix
background}), enabling users to easily extend the parser to deal with new
log lines is a design imperative (\sectionref{parser design}).  Despite the
resulting initial increase in complexity the task is quite tractable,
though it does raise efficiency and optimisation questions (answered in
\sectionref{parser efficiency}).  Providing a tool to ease the generation
of \regexes{} from unparsed log lines should greatly help users add rules
to parse previously unparsed log lines (\sectionref{creating new rules}).

The architecture not only substantially eases the parsing of new log
lines, it makes adding new actions (\sectionref{adding new actions}) a
relatively simple task.  The simplicity of adding a new action frees the
implementor from worrying about how their action might interfere with
existing parsing, so they can focus on correctly implementing their new
action.

The division of the parser into rules, actions, and framework is unusual
because rules are separated so completely from the actions and framework;
although parsers are often divided (the combination of \texttt{lex} and
\texttt{yacc}~\cite{lex-and-yacc} being an obvious example), the parts are
usually quite internally interdependent, and will be combined into a
complete parser by the compilation process; in contrast \parsername{} keeps
the rules and actions separate until the parser runs.  The actions and
framework are not as completely separated, as the actions depend on
services provided by the framework; however the actions and framework are
not tightly integrated: it would be possible, with some work, to completely
separate the two.  XXX I SHOULD DO THAT\@.

% XXX THIS SHOULD PROBABLY BE REMOVED\@.
%The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
%behaviour}) exhibited by the rules and actions is also interesting, and is
%discussed after the flow chart (\sectionref{flow-chart}) and explanation of
%the paths through the parser.  This emergent behaviour greatly eases the
%process of adding new actions (\sectionref{adding new actions}), as the
%actions do not have to be inserted into an explicit flow of control; new
%actions will naturally find their place in the paths through the parser.

The flow of control in this parser is quite different from that of most
other parsers.  In most traditional parsers the parser has a current state,
and each state has a fixed set of acceptable next states, with unacceptable
states causing parsing to fail --- e.g.\ when a \textbf{C} parser sees the
keyword \textbf{for} it expects to immediately see a left parenthesis, with
any other input causing parsing to fail (comments will already have been
removed by the preprocessor).  This parser is different: the rule which
matches the next log line for a mail dictates the action that will be
invoked, which is equivalent to the next state in other parsers; thus the
next log line for a mail dictates the next state for that mail.  XXX THINK
ABOUT HOW TO MAKE THE PARSER MORE STATEFUL IF REQUIRED\@; ADD IT TO THE
ARCHITECTURE SECTION\@.

The real difficulties arise once the parser is successfully dealing with
90\% of the log lines, as the irregularities and complications explained in
\sectionref{complications} only become apparent once the vast majority of
log lines have been parsed successfully.  Adding new rules to deal with
numerous, infrequently occurring log lines is a simple, albeit tiresome,
task, (though alleviated by the tool described in \sectionref{adding new
actions}), whereas dealing with mails where information appears to be
missing is much more grueling.  Trawling through the log files, looking for
something out of the ordinary, possibly hundreds or even thousands of lines
away from the last mention of the queueid in question, is extremely time
consuming and error prone.  Sometimes the task is not to spot the line that
is unusual, but to spot that a line that is normally present is missing,
i.e.\ to realise that one line amongst thousands is absent.  In all cases
the evidence must be used to construct a hypothesis to explain the
irregularities, and that hypothesis must then be tested in the parser; if
successful, the parser must be modified to deal with the irregularities,
without adversely affecting the existing functionality.  The complications
described in \sectionref{complications} were solved in the order they are
described in, and that order closely resembles the frequency with which
they occur; the most frequently occurring complications dominate the
warning messages produced, and so naturally they are the first
complications to be dealt with.

### 2008-10-30 doc/logparser-glossary.tex
First pass through corrections for sections 1-3.
Second pass through corrections for sections 1-3.
Fixed underfull hbox warnings. Dealt with more corrections; nearly finished (if I ignore the 103 XXX notes).

% vim: set filetype=tex :
% The contents of the glossary.
\glossary{name={SQLite3},description={
    \textit{SQLite is a small C library that implements a self-contained,
    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
    \SQL{} implementation focusing on correctness, simplicity and speed.
    Unlike other \SQL{} implementations it does not require a separate
    server process, greatly simplifying deployment of an application
    utilising it.  More details can be found at~\cite{sqlite-features} or
    \url{http://www.sqlite.org/}.
}}

@@

\glossary{name={Bayesian spam filtering},description={
    Bayesian spam filtering is a method of classifying mail based on the
    frequency that the words in the mail have previously appeared in a spam
    corpus and a ham (non-spam) corpus.  A full description is beyond the
    scope of this document, see~\cite{bayesian-filtering, a-plan-for-spam}
    for a detailed explanation.
}}

@@

\glossary{name={hash},description={
    A hashing function transforms a string of characters to a number.
    There are many possible uses for the resulting number: a common usage
    is to maintain a data structure indexed by strings in an efficient
    manner.  A full description is beyond the scope of this paper, further
    information can be found at~\cite{hash-functions}.
}}

>>>>

% vim: set filetype=tex :
% The contents of the glossary.
\glossary{name={SQLite3},description={
    \textit{SQLite is a small C library that implements a self-contained,
    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
    \SQL{} implementation focusing on correctness, simplicity and speed.
    Unlike other \SQL{} implementations it does not require a separate
    server process, greatly simplifying deployment of an application
    utilising it.  More details can be found
    at~\url{http://www.sqlite.org/}.
}}

@@

\glossary{name={Bayesian spam filtering},description={
    Bayesian spam filtering is a method of classifying mail based on the
    frequency that the words in the mail have previously appeared in a spam
    corpus and a ham (non-spam) corpus.  A full description is beyond the
    scope of this thesis, see~\cite{bayesian-filtering,a-plan-for-spam}
    for a detailed explanation.
}}

@@

%\glossary{name={hash},description={
%    A hashing function transforms a string of characters to a number.
%    There are many possible uses for the resulting number: a common usage
%    is to maintain a data structure indexed by strings in an efficient
%    manner.  A full description is beyond the scope of this paper, further
%    information can be found at~\cite{hash-functions}.
%}}

### 2008-10-30 doc/logparser-other-parsers-review.tex
Fixed all overfull hbox warnings, and made some small improvements in the process.
Fixed underfull hbox warnings. Dealt with more corrections; nearly finished (if I ignore the 103 XXX notes).

    Isoqlog is an MTA log analysis program written in C. It designed to
    scan qmail, postfix, sendmail and exim logfile and produce usage
    statistics in \HTML{} format for viewing through a browser. It produces
    Top domains output according to Sender, Receiver, Total mails and
    bytes; it keeps your main domain mail statistics with regard to Days
    Top Domain, Top Users values for per day, per month and years.

@@

    \item The top ten \SMTP{} error codes (XXX SENDING, RECEIVING, OR
        BOTH\@?), e.g.\ the first entry in the list is: \texttt{Requested
        mail action not taken: relaying not allowed, unknown recipient
        user, \ldots}

@@

% Parsed lines in file: 10416129
%  Found 4 dropped records,
%  Found 1176050 corrupted records,
%  Found 0 old records,
%  Found 9240075 new qualified records.

@@

    The Anteater project is a Mail Traffic Analyser. Anteater supports
    currently the logformat produced by Sendmail and by Postfix. The tool
    is written in 100\% C++ and is very easy to customize. Input, output,
    and the analysis are modular class objects with a clear interface.
    There are eight useful analyse modules, writing the result in plain
    ASCII or \HTML{}, to stdout or to files.

@@

YAALA uses a plugin based system to analyse log files and produce \HTML{}
output reports, with all the parsing and report generation handled by
modules.  Using YAALA as a base would be only slightly less work than
starting from scratch, as both input and output modules would need to be
written specially; it may even be more work to implement the parser within
the constraints of YAALA\@.  YAALA supports storing previously gathered
data using Perl's Storable module~\cite{perl-storable}, so other Perl
programs can use Storable to load, examine, and optionally modify the data;
\parsername{} uses a well documented database which is accessible from the
majority of programming languages.  This information was gleaned from the
source code, as the documentation is sadly lacking.

@@

Logrep does not come with any documentation, though some scant
documentation is available on its website (\parsername{} provides copious
documentation).  It requires a web browser to interact with it, so
automated log processing will be difficult, whereas automated processing is
a key part of \parsernames{} design.  Sadly all the author's attempts to
use Logrep failed, as it was unable to access the log files selected; this
appears to be a bug rather than operator error.  If it was caused by
operator error, the interface needs improvement as the (minimal)
instructions were followed as closely as possible, and multiple attempts
were made.  As parsing failed it was not possible to review the reports
Logrep can generate (available in \HTML{} only), or to examine the
(undocumented) format in which it can save extracted data for subsequent
reuse.

>>>>

    Isoqlog is an MTA log analysis program written in C. It designed to
    scan qmail, postfix, sendmail and exim logfile and produce usage
    statistics in HTML format for viewing through a browser. It produces
    Top domains output according to Sender, Receiver, Total mails and
    bytes; it keeps your main domain mail statistics with regard to Days
    Top Domain, Top Users values for per day, per month and years.

@@

    \item A summary of the \SMTP{} codes used when rejecting delivery
        attempts.

@@

XXX NEED A CONCLUSION\@.

@@

    The Anteater project is a Mail Traffic Analyser. Anteater supports
    currently the logformat produced by Sendmail and by Postfix. The tool
    is written in 100\% C++ and is very easy to customize. Input, output,
    and the analysis are modular class objects with a clear interface.
    There are eight useful analyse modules, writing the result in plain
    ASCII or HTML, to stdout or to files.

@@

YAALA uses a plugin based system to analyse log files and produce HTML
output reports, with all the parsing and report generation handled by
modules.  Using YAALA as a base would be only slightly less work than
starting from scratch, as both input and output modules would need to be
written specially; it may even be more work to implement the parser within
the constraints of YAALA\@.  YAALA supports storing previously gathered
data using Perl's Storable module~\cite{perl-storable}, so other Perl
programs can use Storable to load, examine, and optionally modify the data;
\parsername{} uses a well documented database which is accessible from the
majority of programming languages.  This information was gleaned from the
source code, as the documentation is sadly lacking.

@@

Logrep does not come with any documentation, though some scant
documentation is available on its website (\parsername{} provides copious
documentation).  It requires a web browser to interact with it, so
automated log processing will be difficult, whereas automated processing is
a key part of \parsernames{} design.  Sadly all the author's attempts to
use Logrep failed, as it was unable to access the log files selected; this
appears to be a bug rather than operator error.  If it was caused by
operator error, the interface needs improvement as the (minimal)
instructions were followed as closely as possible, and multiple attempts
were made.  As parsing failed it was not possible to review the reports
Logrep can generate (available in HTML only), or to examine the
(undocumented) format in which it can save extracted data for subsequent
reuse.

### 2008-10-30 doc/logparser-results.tex
Fixed underfull hbox warnings. Dealt with more corrections; nearly finished (if I ignore the 103 XXX notes).

@@

@@

    \item Postfix accepts the bounce notification and delivers it to both
        the local mailbox and to username@domain, as instructed by the
        user's forwarding instructions.  A
        \texttt{Delivered-To:~username@domain} header is added to the
        forwarded bounce notification, which now has an envelope sender
        address of username@domain.

    \item Postfix accepts the forwarded bounce notification but during
        deliver the mail it notices that the \texttt{Delivered-To}
        header already contains the address currently being delivered to,
        and sends a bounce notification to the sender: username@domain.

@@

The sequence described above occurs extremely rapidly because Postfix does
not have to deliver the mail to an external system, so mails are delivered,
bounced and generated as fast as the disks can keep up, resulting in a huge
volume of log lines.

The different distribution of log lines during a mail forwarding loop has
unexpected effects on parsing time, described in detail in
\sectionref{rule ordering for efficiency}, \sectionref{scalability as the
number of rules rises}, and \sectionref{Caching compiled regexes}.

@@

\begin{description}

@@

\end{description}

@@

I THINK THIS SECTION NEEDS SUBSTANTIAL WORK, IF NOT A RE-WRITE\@.

>>>>

XXX REUSE THE CONTENT FROM THE PAPER FOR THIS SECTION\@.

@@

XXX REWRITE THIS SECTION SO THAT IT'S CORRECT\@: POINT 4 SAYS THAT THE
ENVELOPE SENDER CHANGES, IS THAT CORRECT\@?

@@

    \item Postfix accepts the bounce notification and delivers it to both
        the local mailbox and to username@domain, as instructed by the
        user's forwarding instructions.  A
        \texttt{Delivered-To:~username@domain} header is added to the
        forwarded bounce notification, XXX which now has an envelope sender
        address of username@domain.   XXX IS THIS CORRECT\@?

    \item Postfix accepts the forwarded bounce notification but during mail
        delivery it notices that the \texttt{Delivered-To} header already
        contains the address currently being delivered to, and sends a
        bounce notification to the sender: username@domain.

@@

The sequence described above occurs extremely rapidly because Postfix does
not have to deliver the mail to an external system, so mails are delivered,
bounced, and generated as fast as the disks can keep up, resulting in a huge
volume of log lines.

The different distribution of log lines caused by a mail forwarding loop
has unexpected effects on parsing time, described in detail in
\sectionref{rule ordering for efficiency}, \sectionref{scalability as the
number of rules rises}, and \sectionref{Caching compiled regexes}.

@@

\begin{eqlist}

@@

\end{eqlist}

@@

XXX I THINK THIS SECTION NEEDS SUBSTANTIAL WORK, IF NOT A RE-WRITE\@; START
WITH THE PAPER'S CONTENT\@.
